---
title: "Lower Bounds on the Convergence Rate of Accept-Reject-Based Markov Chains"
author: "Galin Jones^[Joint work with Austin Brown, Texas A&M]"
institute: "University of Minnesota"
date: "July 2026"
output:  
 beamer_presentation:
# keep_tex: true
 includes:
  in_header: ~/Dropbox/beamer-header-simple.txt
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
require(mcmcse)
require(mvtnorm)
require(SimTools)
```

## Accept-Reject-Based Markov Chains

Any Markov kernel of the form
\[P(x, B) = \int_{B} a(x, y) q(y \mid x) d\lambda(y) + \delta_{x}(B) [1 - A(x)]\]

This includes, among others,

\begin{itemize}
\item[] Metropolis-Hastings

\item[] Non-reversible Metropolis-Hastings

\item[] Barker's

\item[] Portkey Barker's

\item[] Lazy Metropolis-Hastings

\end{itemize}

But the focus will be on Metropolis-Hastings.

## Metropolis-Hastings

Let $f$ be a target density and $q_h$ a proposal density.  

Given $X_t = x$, draw $Y \sim q_{h}(\cdot \mid x)$

Draw $U \sim Unif(0,1)$ and set $X_{t+1} = y$ if 

$$u \le \frac{f(y)q_{h}(x \mid y)}{f(x) q_{h}(y \mid x)}$$

otherwise set $X_{t+1} = x$ \bigskip

Choice of $h$ is crucial to finite-sample performance


## Illustrative Example

$X_1, \ldots, X_n \stackrel{iid}{\sim} N(\theta, 1), ~~~~~~~~~~\theta \sim N(0,1)$

$$\theta \mid x_1, \ldots, x_n \sim N \left( \frac{n \bar{x}_n}{n+1}, \frac{1}{n+1}\right)$$
\bigskip

Use random walk Metropolis-Hastings with Normal proposal having variance $h$



## n=2, h=1

```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}

set.seed(1228)

msim <- 1e3

n <- 2
h <- 1

sim_dat <- rnorm(n, mean = 0, sd = 1)
xbar <- mean(sim_dat)

target_mean <- n*xbar / (n+1)
target_sd <- 1/(1+n)

mh <- matrix(NA_real_, nrow = msim, ncol = 1)

mh[1,] <- 0

acc_rate <- 0

for (iter in 2:msim){
  curr <- mh[iter-1,]
  prop <- rnorm(1, mean = curr, sd = sqrt(h))
  log_acc <- dnorm(prop, mean = target_mean, sd = target_sd, log = TRUE) -               dnorm(curr, mean = target_mean, sd = target_sd, log = TRUE)
  
  if (log_acc >=0) {
    mh[iter,] <- prop
    acc_rate <- acc_rate + 1
    }
  else if (log(runif(1)) <= log_acc){
    mh[iter,] <- prop
    acc_rate <- acc_rate + 1
    }
  else {mh[iter,] <- curr}
}

acc_rate/msim

```

```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
par(mfrow=c(1,2))
ts.plot(mh[], ylab="", xlab = "Iteration", main = "")
acf(mh)
```


## n=20, h=1

```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}

msim <- 1e3

n <- 20

sim_dat <- rnorm(n, mean = 0, sd = 1)
xbar <- mean(sim_dat)

target_mean <- n*xbar / (n+1)
target_sd <- 1/(1+n)

mh <- matrix(NA_real_, nrow = msim, ncol = 1)

mh[1,] <- 0

acc_rate <- 0

for (iter in 2:msim){
  curr <- mh[iter-1,]
  prop <- rnorm(1, mean = curr, sd = sqrt(h))
  log_acc <- dnorm(prop, mean = target_mean, sd = target_sd, log = TRUE) -               dnorm(curr, mean = target_mean, sd = target_sd, log = TRUE)
  
  if (log_acc >=0) {
    mh[iter,] <- prop
    acc_rate <- acc_rate + 1
    }
  else if (log(runif(1)) <= log_acc){
    mh[iter,] <- prop
    acc_rate <- acc_rate + 1
    }
  else {mh[iter,] <- curr}
}

acc_rate/msim

```

```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
par(mfrow=c(1,2))
ts.plot(mh[], ylab="", xlab = "Iteration", main = "")
acf(mh)
```

## n=200, h=1

```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}

msim <- 1e3

n <- 2e2

sim_dat <- rnorm(n, mean = 0, sd = 1)
xbar <- mean(sim_dat)

target_mean <- n*xbar / (n+1)
target_sd <- 1/(1+n)

mh1<- matrix(NA_real_, nrow = msim, ncol = 1)

mh1[1,] <- 0

acc_rate <- 0

for (iter in 2:msim){
  curr <- mh1[iter-1,]
  prop <- rnorm(1, mean = curr, sd = sqrt(h))
  log_acc <- dnorm(prop, mean = target_mean, sd = target_sd, log = TRUE) -               dnorm(curr, mean = target_mean, sd = target_sd, log = TRUE)
  
  if (log_acc >=0) {
    mh1[iter,] <- prop
    acc_rate <- acc_rate + 1
    }
  else if (log(runif(1)) <= log_acc){
    mh1[iter,] <- prop
    acc_rate <- acc_rate + 1
    }
  else {mh1[iter,] <- curr}
}

acc_rate/msim

```

```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
par(mfrow=c(1,2))
ts.plot(mh1[], ylab="", xlab = "Iteration", main = "")
acf(mh1)
```

## n=200, h=1/2000

```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}

msim <- 1e3

n <- 2e2
h <- 1/(10*n)

sim_dat <- rnorm(n, mean = 0, sd = 1)
xbar <- mean(sim_dat)

target_mean <- n*xbar / (n+1)
target_sd <- 1/(1+n)

mh <- matrix(NA_real_, nrow = msim, ncol = 1)

mh[1,] <- 0

acc_rate <- 0

for (iter in 2:msim){
  curr <- mh[iter-1,]
  prop <- rnorm(1, mean = curr, sd = sqrt(h))
  log_acc <- dnorm(prop, mean = target_mean, sd = target_sd, log = TRUE) -               dnorm(curr, mean = target_mean, sd = target_sd, log = TRUE)
  
  if (log_acc >=0) {
    mh[iter,] <- prop
    acc_rate <- acc_rate + 1
    }
  else if (log(runif(1)) <= log_acc){
    mh[iter,] <- prop
    acc_rate <- acc_rate + 1
    }
  else {mh[iter,] <- curr}
}

acc_rate/msim

```

```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
par(mfrow=c(1,2))
ts.plot(mh[], ylab="", xlab = "Iteration", main = "")
acf(mh)
```

## Observations

The proposal scaling needs to account for the sample size and the dimension.

\vskip 0.25in

Significant theoretical efforts through optimal scaling arguments and mixing time analyses. 

\vskip 0.25in

Practitioners largely use trial and error, basing their decisions on plots.

\vskip 0.25in

Can we identify when failure is guaranteed in general MH simulations so that we can avoid it?

## Notation

The MH Markov kernel describes the dynamics of the Markov chain.  Informally,
$$ \Pr (X_{t+j} \in B \mid X_{j} = x) = P^t (x, B)$$
\bigskip

$$ \| P^t (x, \cdot) - F(\cdot) \|_{TV} \to 0 ~~~~ t \to \infty$$
\bigskip

\underline{Question}:  When will this convergence take prohibitively long?

## Lower Bound

\underline{Theorem} If 
$$
A_{h} (x) = \int \left[
\frac{ f(x') q_{h}(x \mid x') }{f(x) q_{h}(x' \mid x)} \wedge 1 \right] q_h(x' \mid x) dx'.
$$
then, for every $x$,
$$ \| P^t(x, \cdot) - F(\cdot) \|_{TV} \ge \left[ 1 - A_{h}(x) \right]^t$$
\bigskip

\underline{Answer}:  Avoid $A_{h}(x) \approx 0$.

## Gaussian Proposals

Suppose $\mu : \mathbb{R}^d \to \mathbb{R}^d$ and consider a proposal of the form
$$ N_d (\mu(x), h C)$$

For example:
\begin{itemize}
\item[-] Independence Sampler: $\mu(x) = c$
\medskip

\item[-] Random Walk: $\mu(x) = x$
\medskip

\item[-] MALA: $\mu(x) = x + h (\nabla \log x)/2$
\end{itemize} 

## Gaussian Proposals

Suppose $\mu : \mathbb{R}^d \to \mathbb{R}^d$ and consider a proposal of the form
$$ N_d (\mu(x), h C),$$
then
$$ \| P^t(x, \cdot) - F(\cdot) \|_{TV} \ge \left[ 1 - \frac{1}{f(x) (2\pi h)^{d/2} \det(C)^{1/2}} \right]^t$$
Suggests 
\begin{itemize}
\item[] $h$ must be small to avoid poor convergence properties

\item[] MH chains can have poor dimension dependence unless that scaling is chosen carefully: 

$h \propto d^{-\delta}$ for some $\delta>0$ 
\end{itemize}


## Geometric Ergodicity

$P$ is geometrically ergodic if there exists $\rho < 1$ such that
$$ \| P^t(x, \cdot) - F(\cdot) \|_{TV} \le M(x) \rho^t $$ 

Key sufficient condition for *reliable* MCMC
\begin{itemize}
\item[-] Markov chain Central Limit Theorem
$$ \sqrt{m}(\bar{X}_{m} - E_F(X)) \to \text{N}_{p}(0, \Sigma)$$
\item[-] Consistency of estimators of $\Sigma$ such as batch means
\end{itemize}

## Geometric Ergodicity

$P$ is geometrically ergodic if there exists $\rho < 1$ such that
$$ \| P^t(x, \cdot) - F(\cdot) \|_{TV} \le M(x) \rho^t $$ 

\vskip 0.1in

Few results on constraining $\rho$ for MH algorithms. That is, there are almost no results that identify constants such that
$$ C_1 \le \rho \le  C_2$$
\vskip 0.1in

Most results on geometric ergodicity of MH establish only the *existence* of $0 < \rho < 1$.

## Illustrative Example n=200, h=1


```{r, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
par(mfrow=c(1,2))
ts.plot(mh1[], ylab="", xlab = "Iteration", main = "")
acf(mh1)
```


## Geometeric Ergodicity

$P$ is geometrically ergodic if there exists $\rho < 1$ such that
$$ \| P^t(x, \cdot) - F(\cdot) \|_{TV} \le M(x) \rho^t $$ 
\bigskip



\underline{Theorem} If $P$ is geometrically ergodic, then
$$ \rho \ge 1 - \inf_{x} A_h(x)$$

## Another Toy Example

Let $b > 1$ and
$$
g(x,y) = \frac{b}{\pi} e^{-(x^2 + b^2 y^2)} \quad \text{ and } \quad h(x,y) = \frac{b}{\pi} e^{-(b^2x^2 + y^2)} .
$$
Set 
$$
f(x,y) = \frac{1}{2} g(x,y) + \frac{1}{2} h(x,y)
$$
RWMH using a Gaussian proposal with scaling $h$ is geometrically ergodic. \bigskip

Our result says
$$  \rho \ge 1 - \frac{1}{2bh}$$

## Bayesian Logistic Regression with Zellner's $g$-prior

$$
f_n(\beta)
\propto \prod_{i = 1}^n s\left( \beta^T X_i \right)^{Y_i} \left( 1 -  s\left( \beta^T X_i \right) \right)^{1 - Y_i} \exp\left( - \frac{1}{2 g} \beta^T X^T X \beta \right).
$$
\bigskip

\underline{Proposition}
Let $\beta_n^*$ denote the point which maximizes $\pi_n$.
If $n \to \infty$ in such a way that $d_n/n \to \gamma \in (0, 1)$, then, with probability 1, for all sufficiently large $n$, the acceptance probability for RWMH satisfies
\[
\| P^{t}(\beta_n^*, \cdot) - F_{n}(\cdot)\|_{TV} \ge \left[1 -  \left( \frac{h n (1 - \sqrt{\gamma})^2}{2 g} + 1 \right)^{-d_n/2}\right]^{t}
.
\]


## Bayesian Logistic Regression with Zellner's $g$-prior

\begin{figure}[t]
\centering
  \includegraphics[width=\linewidth]{acceptance_plot.png}
  \caption{}
\end{figure}


## Bayesian Logistic Regression with a flat prior

Suppose for $i=1, \ldots, n$, $(Y_i, X_i)$ are iid and
$$Y_i \mid X_i, \beta  \stackrel{ind}{\sim} \text{Bern}\left( \left( 1 + \exp\left( -\beta^T X_i \right) \right)^{-1} \right)$$
and $\nu(d\beta) = d\beta$.  

\vskip .25in

\underline{Theorem} If $\beta_n^*$ maximizes the posterior and $d_n \le n^\kappa$, $\kappa \in (0,1)$, the posterior concentrates at $\beta^*_n$, under regularity conditions. 

## Bayesian Logistic Regression with a flat prior

If the proposal is $N_d(\mu(\beta), hC)$, then MH satisfies
$$\| P^{t}(\beta_n^*, \cdot) - F_{n}(\cdot)\|_{TV} \ge \left[ 1 - K \left( \frac{1}{n h} \right)^{d_n/2} \right]^{t} $$
\vskip .25in

\underline{Take Home Message}: Proposal scaling needs to depend carefully on both $n$ and $d$.

## Bayesian Logistic Regression with a flat prior

\begin{figure}[t]
\centering
  \includegraphics[width=\linewidth]{flat_acceptance_plot.png}
\end{figure}


## What's in the paper

\begin{itemize}
\item[] Lower bounds in both total variation and Wasserstein distances 

\item[] Comparison to conductance methods 

\item[] General lower bounds under posterior concentration 

\item[] Existence of a spectral gap is equivalent to a geometric rate of convergence in many Wasserstein distances
\end{itemize}

\vskip .25in

Brown and Jones (2025) Lower Bounds on the Rate of Convergence for Accept-Reject-Based Markov Chains in Wasserstein and Total Variation Distances, \textit{Bernoulli} DOI: 10.3150/24-BEJ1791





